Software architectures to integrate workflow engines in science gateways

### 2016-10-12 paper.tex
Still more answers to the reviewers
Even more answers to the reviewers

\begin{abstract}
  Science gateways often rely on workflow engines to execute
  applications on distributed infrastructures. We investigate six software
  architectures commonly used to integrate workflow engines into
  science gateways. In \emph{tight integration}, the workflow engine
  shares software components with the science gateway. In
  \emph{service invocation}, the engine is isolated and invoked
  through a specific software interface. In \emph{task encapsulation},
  the engine is wrapped as a computing task executed on the
  infrastructure. In the \emph{pool model}, the engine is bundled in
  an agent that connects to a central pool to fetch and execute
  workflows. In \emph{nested workflows}, the engine is integrated as a
  child process of another engine. In \emph{workflow conversion}, the
  engine is integrated through workflow language conversion. We
  describe and evaluate these architectures with metrics for
  assessment of integration complexity, robustness, extensibility,
  scalability and support for meta-workflows and fine-grained
  debugging. Tight integration and task encapsulation are the easiest
  to integrate and the most robust. Extensibility is equivalent in
  most architectures. The pool model is the most scalable one and
  meta-workflows are only available in task encapsulation, nested
  workflows and workflow conversion. These results provide insights
  for science gateway architects and developers.
\end{abstract}

@@

\correction{The reminder of this section provides} background
information and definitions of workflow engines, science gateways and
infrastructures. In Section~\ref{sec:architectures}, we describe six
architectures that cover the main patterns that we encountered in the
various systems that we studied during the last few years. The
architectures are described within a consistent framework that
underlines the functional interactions between their main software
components. \correction{In Section~\ref{sec:metrics}, we introduce
  metrics that measure integration complexity, robustness,
  extensibility, scalability and other specific features. These
  metrics are specifically designed to measure the ability of the
  architectures to address workflow-related issues commonly
  encountered in science gateways. In Section~\ref{sec:evaluation}, we
  evaluate the different architectures based on the proposed metrics.}
We finally compare the architectures and highlight recommendations for
science gateway architects and developers.

@@

% Software properties: integration effort, extensibility, robustness,
% scalability, specific features

@@

Architectures to integrate workflow engines in science gateways are
presented in Figure~\ref{fig:architectures} using the graphical
notations defined in
Figure~\ref{fig:notations}. Table~\ref{table:system-classification}
summarizes the classification of a few systems by architecture.

@@

This architecture is largely adopted in systems such as Apache
Airvata~\cite{marru2011apache}, GridSFEA~\cite{Elts2010}, Vine
Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual
Imaging Platform~\cite{GLAT-13}, the systems
in~\cite{wu2010accelerating,shahand:2012jgc} and the WS-PGRADE/gUSE
framework~\cite{Kacsuk2012}. The integration between the HubZero
science gateway and the Pegasus workflow engine performed
in~\cite{CPE:CPE3257} also uses a service architecture, where the
\texttt{d} interaction is implemented as a set of calls to Pegasus'
command-line tools (e.g., pegasus-status, pegasus-plan,
etc.). Figure~\ref{fig:vip-architecture} shows the architecture of the
Virtual Imaging Platform, where the MOTEUR workflow
engine~\cite{GLAT-08i} is invoked through a service.

@@

The science gateway also transfers both user and workflow data across
the infrastructure, through interaction \texttt{c$_1$}.

@@

\subsection{Pool model}
\label{sec:pool}
See Figure~\subref*{archi:agent}. Workflows are submitted by the
science gateway to a central pool through interaction
\texttt{d}. Agents connect to the pool asynchronously to retrieve and
execute workflows through interaction \texttt{f}. Agents may be
started according to various policies, for instance to ensure load
balancing. Workflow engine controls tasks and data on the
infrastructure through interactions \texttt{b} and \texttt{c$_2$},
science gateway transfers user data through interaction
\texttt{c$_1$}, and administrator installs workflows through
interaction \texttt{a$_1$} and \texttt{a$_2$}.

@@

Nested workflows have long been available in workflow engines, for
instance in the Taverna workbench~\cite{oinn2004taverna}. They are
also used implicitly in several platforms where workflow engines are
wrapped in workflow tasks as any other command-line tool. Nested
workflows were notably used by the SHIWA Science Gateway to implement
so-called Coarse-Grained workflow
interoperability~\cite{terstyanszky2014enabling}, i.e. to integrate
various workflow engines in a consistent
platform. Figure~\ref{fig:shiwa-architecture} shows the architecture
used in the SHIWA Simulation Platform for nested workflow execution
with service invocation.
\begin{figure*}
\centering
\includegraphics[width=1.5\columnwidth]{figures/shiwa-science-gateway.pdf}
\caption{Nested workflow execution through the SHIWA Science Gateway. The
  parent workflow engine is \texttt{WS-PGRADE/gUSE}, invoked as a service
  in the Science Gateway (step \texttt{1}, interaction \texttt{d} on
  Figure~\ref{fig:architectures}(V)-Right). Ten different child engines can be used by nested
  workflows, invoked through the \texttt{Submission service} (step
  \texttt{2}, interaction \texttt{d}). Each of these engines can
  submit tasks and transfer data to a distributed computing
  infrastructure (\texttt{DCI}, step \texttt{3}, interactions
  \texttt{b} and \texttt{c$_2$}). User data interactions (\texttt{c$_1$}) and
  application porting ones (\texttt{a}) are not represented. Figure
  reproduced from~\cite{terstyanszky2014enabling} with permission of
  the first author.}
\label{fig:shiwa-architecture}
\end{figure*}

@@

The architectures described in Section~\ref{sec:architectures} are
evaluated in Table~\ref{table:evaluation} using five main criteria:
integration complexity, robustness, extensibility, scalability and
other specific features. These are based on science gateway properties
introduced in Section~\ref{sec:intro_sg}. Criteria break down into
specific metrics where \emph{lower value indicates better
  performance}. Colors in Table~\ref{table:evaluation} represent
normalized scores (see caption). For each criterion, a global score is
computed by summing up the individual scores for each metric. We
ensure that the different metrics used to calculate the global score
for a criterion are comparable, so that they can be combined
meaningfully. \correction{All the metrics in a criterion contribute
  equally to the global score. If this is not the case, the
  reader is invited to adjust the weights in the public spreadsheet available
  at \url{https://frama.link/LrPRsCQp} (column B of sheet ``Metrics'')
  to generate custom versions of Table~\ref{table:evaluation}.}

@@

\cellcolor[HTML]{EEEEEE}\textbf{Specific features}& \multicolumn{6}{l}{\cellcolor[HTML]{EEEEEE}}\\
  Meta-workflow  -- \texttt{O$_1$}    & \cellcolor[HTML]{999999}1

@@

  Fine-grained debugging -- \texttt{O$_2$}   & \cellcolor[HTML]{99FF99}0

@@

@@

\emph{Robustness of workflow execution} measures the likelihood that the
workflow execution fails due to errors in the components or in the
interactions of the software architecture. Errors coming
from the infrastructure (e.g., unavailable data or terminated tasks)
or workflows (e.g., wrong user input or application errors) are not
covered, since they do not stem directly from the software
architecture. 
Robustness is measured here as a consequence of global
complexity, since complex architectures tend to be more prone to
failure. More precisely, robustness is determined by the number of
software components and interactions represented in
red in the architecture diagrams in
Figure~\ref{fig:architectures}, which are specific to workflow
execution. Two metrics are used:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Specific components (\texttt{R$_1$}): number of specific
  components involved in workflow executions. \correction{For instance, the
  workflow service is a specific component because its main objective
  is to execute workflows. On the contrary, the science gateway is
  \emph{not} specific to workflow execution since it is used for a
  variety of other functions such as user authentication, data
  transfer, etc.}
\item Specific interactions (\texttt{R$_2$}): number of specific
  interactions involved in workflow executions. \correction{For
    instance, the pool-agent interaction (interaction \texttt{f}) is
    specific to workflow execution while data control (interaction
    \texttt{c$_1$}) is \emph{not} because it is used by the science gateway to transfer
    user data regardless of a particular workflow execution}.
\end{itemize}
\texttt{R$_1$} and \texttt{R$_2$} are
summed to obtain the global score for this criterion, which measures
the total number of software pieces that are specifically involved in
the workflow execution. \correction{The weights given to software components in R$_1$ and interactions in R$_2$ can be adjusted in the public spreadsheet.}

@@

\emph{Specific features} include:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Meta-workflow (\texttt{O$_1$}): measures the ability to describe
  meta-workflows from existing workflows. Meta-workflows offer an
  additional level of flexibility to build workflows from reusable
  components. \texttt{O$_1$}=0 means that meta-workflows are
  intrinsically enabled by the architecture (i.e. they can be
  implemented using the components and interactions already in place),
  and \texttt{O$_1$}=1 means that they may be implemented with some
  development effort.

\item Fine-grained debugging (\texttt{O$_2$}): availability of
  fine-grained debugging information about workflow tasks (white-box
  workflow). \correction{We call ``white-box'' workflow a workflow
    where detailed information about individual tasks is available,
    for instance task statuses, execution logs and resource
    consumption. On the contrary, a ``black-box'' workflow only
    displays global information such as the workflow status and the
    logs of the workflow engine.} Fine-grained information about
  workflow tasks is required to properly troubleshoot workflow
  executions. \correction{It helps workflow users (scientists)
    identify errors, it allows gateway administrators to troubleshoot
    executions, and it facilitates debugging by workflow
    developers}. \correction{The availability and accuracy of
    fine-grained debugging information} usually depends on the number
  of software layers between the science gateway and the workflow
  engine~\cite{olabarriaga2014}. \texttt{O$_2$}=0 means that the
  information is directly accessible in the science gateway,
  \texttt{O$_2$}=1 means that it is obtained through one or more
  software interactions.
\end{itemize}
\texttt{O$_1$} and \texttt{O$_2$} are summed to obtain a total number of missing specific
features in the architecture.

@@

\paragraph{Specific features} Supporting meta-workflows requires
specific developments in the workflow engine or science gateway to
allow workflows to execute other workflows (\texttt{O$_1$}=1). For
instance, specific Java objects may be implemented in the science
gateway framework to execute different types of workflows and link
them together.  The science gateway can
retrieve fine-grained debugging information from the workflow engine
directly (\texttt{O$_2$}=0).

@@

\paragraph{Specific features} Meta-workflows require specific
developments in the workflow engine or science gateway to allow
workflows to execute other workflows.  (\texttt{O$_1$}=1).
 The
science gateway needs to invoke interaction \texttt{d} to retrieve
fine-grained debugging information from the workflow service
(\texttt{O$_2$}=1).

@@

\paragraph{Specific features} Task encapsulation \correction{does not enable meta-workflows by default (\texttt{O$_1$=1})}. Fine-grained debugging
information is obtained through interaction \texttt{b}
(\texttt{O$_2$}=1).

@@

\paragraph{Specific features} Meta-workflows require specific
developments in the workflow engine to enable workflow submission
(\texttt{O$_1$=1}). Debugging information is accessed through
interactions \texttt{d} and \texttt{f} (\texttt{O$_2$=1}).

@@

\paragraph{Specific features}
Meta-workflows can be implemented, which is one of the main interest
of this architecture (\texttt{O$_1$=0}). Note, however, that the
complexity of the architecture increases with the number of engine
types involved in meta-workflows: for instance, a meta-workflow with
child workflows executed by 2 different types of engines would require an
additional workflow service and the corresponding interactions.
Debugging information is accessed through interaction \texttt{d} (invoked twice) (\texttt{O$_2$=1}).

@@

\paragraph{Specific features} Meta-workflows are available after
conversion, by connecting workflows in the language used in the
science gateway (\texttt{O$_1$=0}).  Debugging information is accessible
as in the service invocation architecture (\texttt{O$_2$=1}).

\correction{\section{Real systems}}

@@

We find that the studied system supports nested workflows with a much
better performance than the nested workflow model instantiated with
service invocation (evaluated on Table~\ref{table:evaluation}): the
integration score is improved by a factor of 2, the robustness is
improved by a factor of 2.6, the extensibility is substantially
improved, the scalability and specific features are not penalized.

@@

\begin{table}
\footnotesize
\centering
\begin{tabular}{rcc}
                                     & \textbf{Nested} & \textbf{Nested}\\
                                     & \textbf{Tight+Task} & \textbf{Task+Task}\\
\cellcolor[HTML]{EEEEEE}\textbf{Integration complexity}&\cellcolor[HTML]{EEEEEE}&\cellcolor[HTML]{EEEEEE}\\
  Total components -- \texttt{I$_1$} & 2 & 2 \\
Total interactions -- \texttt{I$_2$} & 6 & 5 \\
\textbf{Total}& \textbf{8} & \textbf{7}\\
\cellcolor[HTML]{EEEEEE}\textbf{Robustness}&\cellcolor[HTML]{EEEEEE}&\cellcolor[HTML]{EEEEEE}\\
Specific components -- \texttt{R$_1$} & 0 & 0\\
  Specific interactions -- \texttt{R$_2$} & 3 & 1\\
  \textbf{Total}& \textbf{3} & \textbf{1}\\
\cellcolor[HTML]{EEEEEE}\textbf{Extensibility}&\cellcolor[HTML]{EEEEEE}&\cellcolor[HTML]{EEEEEE}\\
  New engine type -- \texttt{E$_1$}  & 3 & 2\\
New engine version -- \texttt{E$_2$} & 1 & 1\\
  New workflow -- \texttt{E$_3$}     & 2 & 2\\
New infrastructure -- \texttt{E$_4$} & 4 & 3\\
  \textbf{Total}& \textbf{10} & \textbf{8}\\
\cellcolor[HTML]{EEEEEE}\textbf{Scalability}& \cellcolor[HTML]{EEEEEE}&\cellcolor[HTML]{EEEEEE}\\
Multiple engine instances -- \texttt{S$_1$}& 1 & 0\\
Distributed engines -- \texttt{S$_2$}& 0 & 0\\
Task scheduling -- \texttt{S$_3$}    & 1 & 1\\
\textbf{Total}& \textbf{2} & \textbf{1}\\
\cellcolor[HTML]{EEEEEE}\textbf{Specific features}&\cellcolor[HTML]{EEEEEE}&\cellcolor[HTML]{EEEEEE}\\
  Meta-workflow  -- \texttt{O$_1$}    & 0 & 0\\
  Fine-grained debugging -- \texttt{O$_2$}  & 1 & 1\\
  \textbf{Total}& \textbf{1} & \textbf{1}\\
\end{tabular}
\caption{\correction{Evaluation of the nested workflow architecture instantiated with tight integration (parent engine) and task encapsulation (child engine).}}
\label{table:evaluation-2}
\end{table}

@@

Differences in other specific features should not be neglected. Task encapsulation, nested
workflows (with service invocation) and workflow conversion (with service
invocation) are the only architectures that intrinsically support meta-workflows. Besides,
tight integration is the only architecture that allows
fine-grained debugging, which may be
critical for efficient user support.

@@

Note also that all the interactions involved in the architectures were
treated equally, whereas some of them are obviously more complex than
others. For example, interaction \texttt{g} (workflow language
conversion) is clearly more complex than interaction \texttt{d}
(service invocation). However, without entering into the details of a particular
system, quantification of the robustness or the amount of development
associated with each interaction and software component will hardly be
precise.

@@

The particular technical or historical context of a science gateway
project may also influence the choice of an architecture to
integrate workflow engines. For instance, many workflow engines are
already available as web services, which tends to favor service
invocation (\correction{in particular when interactions can be re-purposed}), and other science gateways may have strongly tested task
and data control features (interactions \texttt{b} and \correction{\texttt{c$_1$}}),
which would favor task encapsulation. Similarly, adding a new type of engine
may facilitate integration of a new infrastructure when interactions
\texttt{b} and \texttt{c$_2$} are already available. The migration
cost between architectures has been ignored as well.

@@

This work has been made possible with the support of the Irving Ludmer
Family Foundation and the Ludmer Centre for Neuroinformatics and
Mental Health. The integration between PSOM and CBRAIN was supported
by a Brain Canada Platform Support Grant, as well as the Canadian
Consortium on Neurodegeneration in Aging (CCNA), through a grant from
the Canadian Institute of Health Research and funding from several
partners.  This work is in the scope of the LABEX PRIMES (ANR-11-
LABX-0063) of Universit\'e de Lyon, within the program
``Investissements d'Avenir'' (ANR-11-IDEX-0007) operated by the French
National Research Agency (ANR). This work also falls into the scope of
the scientific topics of the French National Grid Institute (IdG). The
VIP team thanks the site administrators of the European Grid
Initiative and the GGUS support for their help related to the VIP
platform. The CBRAIN team is grateful for the computing cycles,
storage, and support obtained from Compute Canada
(\url{https://computecanada.ca}) and platform development program from
CANARIE (\url{http://www.canarie.ca/}). We also acknowledge the Dutch
national e-Infrastructure with the support of SURF Cooperative, the
Dutch national program COMMIT/ and the High Performance Computing and
Networking (HPCN) Fund of the University of Amsterdam for their
support to the science gateway actitivies at the AMC.  We are also
grateful to the financial support provided by FP7 E-INFRASTRUCTURE
program for financial support to SCI-BUS, SHIWA and ER-flow
projects.

>>>>

\begin{abstract}
  Science gateways often rely on workflow engines to execute
  applications on distributed infrastructures. We investigate six software
  architectures commonly used to integrate workflow engines into
  science gateways. In \emph{tight integration}, the workflow engine
  shares software components with the science gateway. In
  \emph{service invocation}, the engine is isolated and invoked
  through a specific software interface. In \emph{task encapsulation},
  the engine is wrapped as a computing task executed on the
  infrastructure. In the \emph{pool model}, the engine is bundled in
  an agent that connects to a central pool to fetch and execute
  workflows. In \emph{nested workflows}, the engine is integrated as a
  child process of another engine. In \emph{workflow conversion}, the
  engine is integrated through workflow language conversion. We
  describe and evaluate these architectures with metrics for
  assessment of integration complexity, robustness, extensibility,
  scalability and \correction{functionality}. Tight integration and task encapsulation are the easiest
  to integrate and the most robust. Extensibility is equivalent in
  most architectures. The pool model is the most scalable one and
  meta-workflows are only available in task encapsulation, nested
  workflows and workflow conversion. These results provide insights
  for science gateway architects and developers.
\end{abstract}

@@

\correction{The reminder of this section provides} background
information and definitions of workflow engines, science gateways and
infrastructures. In Section~\ref{sec:architectures}, we describe six
architectures that cover the main patterns that we encountered in the
various systems that we studied during the last few years. The
architectures are described within a consistent framework that
underlines the functional interactions between their main software
components. \correction{In Section~\ref{sec:metrics}, we introduce
  metrics that measure integration complexity, robustness,
  extensibility, scalability and \correction{functionality}. These
  metrics are specifically designed to measure the ability of the
  architectures to address workflow-related issues commonly
  encountered in science gateways. In Section~\ref{sec:evaluation}, we
  evaluate the different architectures based on the proposed metrics.}
We finally compare the architectures and highlight recommendations for
science gateway architects and developers.

@@

@@

Architectures to integrate workflow engines in science gateways are
presented in Figure~\ref{fig:architectures} using the graphical
notations defined in Figure~\ref{fig:notations}. \correction{Software
  components are defined at a granularity such that two components may
  potentially be operated by different teams on different hardware
  systems. For instance, the science gateway and infrastructure are
  usually distinct components. In real systems, software components
  may exist at a finer granularity, to implement functions such as
  authentication, data persistency or logging. Such fine-grained
  components are not covered by our analysis since we focus on the
  interactions between the science gateway, the infrastructure and the
  workflow engine.}  Table~\ref{table:system-classification}
summarizes the classification of a few systems by architecture.

@@

This architecture is largely adopted in systems such as Apache
Airvata~\cite{marru2011apache}, GridSFEA~\cite{Elts2010}, Vine
Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual
Imaging Platform~\cite{GLAT-13}, the systems
in~\cite{wu2010accelerating,shahand:2012jgc} and the WS-PGRADE/gUSE
framework~\cite{Kacsuk2012}. The integration between the HubZero
science gateway and the Pegasus workflow engine performed
in~\cite{CPE:CPE3257} also uses a service architecture, where the
\texttt{d} interaction is implemented as a set of calls to Pegasus'
command-line tools (e.g., pegasus-status, pegasus-plan,
etc.).

@@

The science gateway also transfers both user and workflow data across
the infrastructure, through interaction \texttt{c$_1$}.
\subsection{Pool model}
\label{sec:pool}
See Figure~\subref*{archi:agent}. Workflows are submitted by the
science gateway to a central pool through interaction
\texttt{d}. Agents connect to the pool asynchronously to retrieve and
execute workflows through interaction \texttt{f}. Agents may be
started according to various policies, for instance to ensure load
balancing. Workflow engine controls tasks and data on the
infrastructure through interactions \texttt{b} and \texttt{c$_2$},
science gateway transfers user data through interaction
\texttt{c$_1$}, and administrator installs workflows through
interaction \texttt{a$_1$} and \texttt{a$_2$}.

Nested workflows have long been available in workflow engines, for
instance in the Taverna workbench~\cite{oinn2004taverna}. They are
also used implicitly in several platforms where workflow engines are
wrapped in workflow tasks as any other command-line tool. Nested
workflows were notably used by the SHIWA Science Gateway to implement
so-called Coarse-Grained workflow
interoperability~\cite{terstyanszky2014enabling}, i.e. to integrate
various workflow engines in a consistent
platform.

\correction{\section{Architectural patterns in real systems}}

%\todo{Introduce this section and take the text out of the captions}

Figure~\ref{fig:vip-architecture} shows the architecture of the
Virtual Imaging Platform, where the MOTEUR workflow
engine~\cite{GLAT-08i} is invoked through a service.

@@

@@

Figure~\ref{fig:shiwa-architecture} shows the architecture
used in the SHIWA Simulation Platform for nested workflow execution
with service invocation.
\begin{figure*}
\centering
\includegraphics[width=1.5\columnwidth]{figures/shiwa-science-gateway.pdf}
\caption{Nested workflow execution through the SHIWA Science Gateway. The
  parent workflow engine is \texttt{WS-PGRADE/gUSE}, invoked as a service
  in the Science Gateway (step \texttt{1}, interaction \texttt{d} on
  Figure~\ref{fig:architectures}(V)-Right). Ten different child engines can be used by nested
  workflows, invoked through the \texttt{Submission service} (step
  \texttt{2}, interaction \texttt{d}). Each of these engines can
  submit tasks and transfer data to a distributed computing
  infrastructure (\texttt{DCI}, step \texttt{3}, interactions
  \texttt{b} and \texttt{c$_2$}). User data interactions (\texttt{c$_1$}) and
  application porting ones (\texttt{a}) are not represented. Figure
  reproduced from~\cite{terstyanszky2014enabling} with permission of
  the first author.}
\label{fig:shiwa-architecture}
\end{figure*}

@@

The architectures described in Section~\ref{sec:architectures} are
evaluated in Table~\ref{table:evaluation} using five main criteria:
integration complexity, robustness, extensibility, scalability and
\correction{functionality}. These are based on science gateway properties
introduced in Section~\ref{sec:intro_sg}. Criteria break down into
specific metrics where \emph{lower value indicates better
  performance}. Colors in Table~\ref{table:evaluation} represent
normalized scores (see caption). For each criterion, a global score is
computed by summing up the individual scores for each metric. We
ensure that the different metrics used to calculate the global score
for a criterion are comparable, so that they can be combined
meaningfully. \correction{All the metrics in a criterion contribute
  equally to the global score. Should the metrics be weighted
  differently in a particular system, the reader is invited to produce
  his/her own version of the Table using the public spreadsheet
  available at \url{https://frama.link/LrPRsCQp} (see column B of
  sheet ``Metrics'').}

@@

\cellcolor[HTML]{EEEEEE}\textbf{\correction{Functionality}}& \multicolumn{6}{l}{\cellcolor[HTML]{EEEEEE}}\\
  Meta-workflow  -- \texttt{\correction{F}$_1$}    & \cellcolor[HTML]{999999}1

@@

  Fine-grained debugging -- \texttt{\correction{F}$_2$}   & \cellcolor[HTML]{99FF99}0

@@

                                     & \cellcolor[HTML]{99FF99}\textbf{\correction{2}}

@@

\emph{Robustness of workflow execution} measures the likelihood that
the workflow execution fails due to errors in the components or in the
interactions of the software architecture. Errors coming from the
infrastructure (e.g., unavailable data or terminated tasks) or
workflows (e.g., wrong user input or application errors) are not
covered, since they do not stem directly from the software
architecture.  Robustness is measured here as a consequence of global
complexity, since complex architectures tend to be more prone to
failure, \correction{in particular in distributed systems where
  interactions between components involve wide-area network
  communications between machines that may be operated by different
  people}. More precisely, robustness is determined by the number of
software components and interactions represented in red in the
architecture diagrams in Figure~\ref{fig:architectures}, which are
specific to workflow execution. Two metrics are used:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Specific components (\texttt{R$_1$}): number of specific
  components involved in workflow executions. \correction{For instance, the
  workflow service is a specific component because its main objective
  is to execute workflows. On the contrary, the science gateway is
  \emph{not} specific to workflow execution since it is used for a
  variety of other functions such as user authentication, data
  transfer, etc.}
\item Specific interactions (\texttt{R$_2$}): number of specific
  interactions involved in workflow executions. \correction{For
    instance, the pool-agent interaction (interaction \texttt{f}) is
    specific to workflow execution while data control (interaction
    \texttt{c$_1$}) is \emph{not} because it is used by the science gateway to transfer
    user data regardless of a particular workflow execution}.
\end{itemize}
\texttt{R$_1$} and \texttt{R$_2$} are
summed to obtain the global score for this criterion, which measures
the total number of software pieces that are specifically involved in
the workflow execution. \correction{The weights given to software components in R$_1$ and interactions in R$_2$ can be adjusted in the public spreadsheet.}

@@

\correction{\emph{Functionality} includes}:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Meta-workflow (\texttt{\correction{F}$_1$}): measures the ability to describe
  meta-workflows from existing workflows. Meta-workflows offer an
  additional level of flexibility to build workflows from reusable
  components. \texttt{\correction{F}$_1$}=0 means that meta-workflows are
  intrinsically enabled by the architecture (i.e. they can be
  implemented using the components and interactions already in place),
  and \texttt{\correction{F}$_1$}=1 means that they may be implemented with some
  development effort.

\item Fine-grained debugging (\texttt{\correction{F}$_2$}): availability of
  fine-grained debugging information about workflow tasks (white-box
  workflow). \correction{We call ``white-box'' workflow a workflow
    where detailed information about individual tasks is available,
    for instance task statuses, execution logs and resource
    consumption. On the contrary, a ``black-box'' workflow only
    displays global information such as the workflow status and the
    logs of the workflow engine.} Fine-grained information about
  workflow tasks is required to properly troubleshoot workflow
  executions. \correction{It helps workflow users (scientists)
    identify errors, it allows gateway administrators to troubleshoot
    executions, and it facilitates debugging by workflow
    developers}. \correction{The availability and accuracy of
    fine-grained debugging information} usually depends on the number
  of software layers between the science gateway and the workflow
  engine~\cite{olabarriaga2014}. \texttt{\correction{F}$_2$}=0 means that the
  information is directly accessible in the science gateway,
  \texttt{\correction{F}$_2$}=1 means that it is obtained through one or more
  software interactions.
\end{itemize}
\texttt{\correction{F}$_1$} and \texttt{\correction{F}$_2$} are summed to obtain a total number of missing 
features in the architecture.

@@

\paragraph{\correction{Functionality}} Supporting meta-workflows requires
specific developments in the workflow engine or science gateway to
allow workflows to execute other workflows (\texttt{\correction{F}$_1$}=1). For
instance, specific Java objects may be implemented in the science
gateway framework to execute different types of workflows and link
them together.  The science gateway can
retrieve fine-grained debugging information from the workflow engine
directly (\texttt{\correction{F}$_2$}=0).

@@

\paragraph{\correction{Functionality}} Meta-workflows require specific
developments in the workflow engine or science gateway to allow
workflows to execute other workflows.  (\texttt{\correction{F}$_1$}=1).
 The
science gateway needs to invoke interaction \texttt{d} to retrieve
fine-grained debugging information from the workflow service
(\texttt{\correction{F}$_2$}=1).

@@

\paragraph{\correction{Functionality}} Task encapsulation
\correction{does not enable meta-workflows by default
  (\texttt{\correction{F}$_1$=1})}. Fine-grained debugging information is obtained
through interaction \texttt{b} (\texttt{\correction{F}$_2$}=1).

@@

\paragraph{\correction{Functionality}} Meta-workflows require specific
developments in the workflow engine to enable workflow submission
(\texttt{\correction{F}$_1$=1}). Debugging information is accessed through
interactions \texttt{d} and \texttt{f} (\texttt{\correction{F}$_2$=1}).

@@

\paragraph{\correction{Functionality}}
Meta-workflows can be implemented, which is one of the main interest
of this architecture (\texttt{\correction{F}$_1$=0}). Note, however, that the
complexity of the architecture increases with the number of engine
types involved in meta-workflows: for instance, a meta-workflow with
child workflows executed by 2 different types of engines would require
an additional workflow service and the corresponding interactions.
Debugging information is accessed through interaction \texttt{d}
(invoked twice) (\texttt{\correction{F}$_2$=1}).

@@

\paragraph{\correction{Functionality}} Meta-workflows are available after
conversion, by connecting workflows in the language used in the
science gateway (\texttt{\correction{F}$_1$=0}).  Debugging information is accessible
as in the service invocation architecture (\texttt{\correction{F}$_2$=1}).

@@

We find that the studied system supports nested workflows with a much
better performance than the nested workflow model instantiated with
service invocation (evaluated on Table~\ref{table:evaluation}): the
integration score is improved by a factor of 2, robustness is
improved by a factor of 2.6, extensibility is substantially
improved, scalability and functionality are not penalized.

@@

\begin{table}
\footnotesize
\centering
\begin{tabular}{rcc}
                                     & \textbf{Nested} & \textbf{Nested}\\
                                     & \textbf{Tight+Task} & \textbf{Task+Task}\\
\cellcolor[HTML]{EEEEEE}\textbf{Integration complexity}&\cellcolor[HTML]{EEEEEE}&\cellcolor[HTML]{EEEEEE}\\
  Total components -- \texttt{I$_1$} & 2 & 2 \\
Total interactions -- \texttt{I$_2$} & 6 & 5 \\
\textbf{Total}& \textbf{8} & \textbf{7}\\
\cellcolor[HTML]{EEEEEE}\textbf{Robustness}&\cellcolor[HTML]{EEEEEE}&\cellcolor[HTML]{EEEEEE}\\
Specific components -- \texttt{R$_1$} & 0 & 0\\
  Specific interactions -- \texttt{R$_2$} & 3 & 1\\
  \textbf{Total}& \textbf{3} & \textbf{1}\\
\cellcolor[HTML]{EEEEEE}\textbf{Extensibility}&\cellcolor[HTML]{EEEEEE}&\cellcolor[HTML]{EEEEEE}\\
  New engine type -- \texttt{E$_1$}  & 3 & 2\\
New engine version -- \texttt{E$_2$} & 1 & 1\\
  New workflow -- \texttt{E$_3$}     & 2 & 2\\
New infrastructure -- \texttt{E$_4$} & 4 & 3\\
  \textbf{Total}& \textbf{10} & \textbf{8}\\
\cellcolor[HTML]{EEEEEE}\textbf{Scalability}& \cellcolor[HTML]{EEEEEE}&\cellcolor[HTML]{EEEEEE}\\
Multiple engine instances -- \texttt{S$_1$}& 1 & 0\\
Distributed engines -- \texttt{S$_2$}& 0 & 0\\
Task scheduling -- \texttt{S$_3$}    & 1 & 1\\
\textbf{Total}& \textbf{2} & \textbf{1}\\
\cellcolor[HTML]{EEEEEE}\textbf{\correction{Functionality}}&\cellcolor[HTML]{EEEEEE}&\cellcolor[HTML]{EEEEEE}\\
  Meta-workflow  -- \texttt{\correction{F}$_1$}    & 0 & 0\\
  Fine-grained debugging -- \texttt{\correction{F}$_2$}  & 1 & 1\\
  \textbf{Total}& \textbf{1} & \textbf{1}\\
\end{tabular}
\caption{\correction{Evaluation of the nested workflow architecture instantiated with tight integration (parent engine) and task encapsulation (child engine).}}
\label{table:evaluation-2}
\end{table}

@@

Differences in \correction{functionality} should not be neglected. Task
encapsulation, nested workflows (with service invocation) and workflow
conversion (with service invocation) are the only architectures that
intrinsically support meta-workflows. Besides, tight integration is
the only architecture that allows fine-grained debugging, which may be
critical for efficient user support.

@@

Note also that all the interactions involved in the architectures were
treated equally, whereas some of them are obviously more complex than
others. For example, interaction \texttt{g} (workflow language
conversion) is clearly more complex than interaction \texttt{d}
(service invocation). However, without entering into the details of a particular \todo{refer to spreadsheet again}
system, quantification of the robustness or the amount of development
associated with each interaction and software component will hardly be
precise.

@@

The particular technical or historical context of a science gateway
project may also influence the \correction{evaluation} of an
architecture to integrate workflow engines. For instance, many
workflow engines are already available as web services, which tends to
favor service invocation (\correction{in particular when interactions
  can be re-purposed}), and other science gateways may have strongly
tested task and data control features (interactions \texttt{b} and
\correction{\texttt{c$_1$}}), which would favor task
encapsulation. Similarly, adding a new type of engine may facilitate
integration of a new infrastructure when interactions \texttt{b} and
\texttt{c$_2$} are already available. The migration cost between
architectures has been ignored as well.

@@

\correction{We thank the anonymous reviewers for the thorough reviews and useful comments that greatly contributed to improve the quality of this paper.} This work has been made possible with the support of the Irving Ludmer
Family Foundation and the Ludmer Centre for Neuroinformatics and
Mental Health. The integration between PSOM and CBRAIN was supported
by a Brain Canada Platform Support Grant, as well as the Canadian
Consortium on Neurodegeneration in Aging (CCNA), through a grant from
the Canadian Institute of Health Research and funding from several
partners.  This work is in the scope of the LABEX PRIMES (ANR-11-
LABX-0063) of Universit\'e de Lyon, within the program
``Investissements d'Avenir'' (ANR-11-IDEX-0007) operated by the French
National Research Agency (ANR). This work also falls into the scope of
the scientific topics of the French National Grid Institute (IdG). The
VIP team thanks the site administrators of the European Grid
Initiative and the GGUS support for their help related to the VIP
platform. The CBRAIN team is grateful for the computing cycles,
storage, and support obtained from Compute Canada
(\url{https://computecanada.ca}) and platform development program from
CANARIE (\url{http://www.canarie.ca/}). We also acknowledge the Dutch
national e-Infrastructure with the support of SURF Cooperative, the
Dutch national program COMMIT/ and the High Performance Computing and
Networking (HPCN) Fund of the University of Amsterdam for their
support to the science gateway actitivies at the AMC.  We are also
grateful to the financial support provided by FP7 E-INFRASTRUCTURE
program for financial support to SCI-BUS, SHIWA and ER-flow
projects.
